---
title: "MRR Project - Genes & Breast Cancer - Introduction"
author: "Julien EYRAUD & Alexandre GOIGNARD"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    #toc: true
    #toc_depth: 3
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
graphics.off()

#remove code parts
knitr::opts_chunk$set(echo = FALSE)
```

## Classification of Cancer outcome using Genetic and Clinical data

# Introduction

This project investigates the effect of genetic and clinical variables on the survival outcome of breast cancer patients. The data contains records of more than 1000 breast cancer patients from several research institutions. Clinical data contains patient-related and tumor-related information. Additionally mRNA gene expression data is available for each patient. The gene expression data has been processed to include only the top 5000 most variable genes on the transformed scale (log2(counts + 1)).

We consider that the main outcome variable of interest is **vital_status**, which is defined in clinical data.

## Load data

```{r cars}
load("mrr_bio.Rdata")

# genetic data: n x p matrix
cat("GeneX dimensions :", dim(GeneX))

# clinical data
cat("Clinical data dimensions :", dim(clinical_data))

table(clinical_data$vital_status)
```

```{r}
clinical_data <- as.data.frame(clinical_data)
GeneX <- as.data.frame(GeneX)

table(unlist(clinical_data$primary_site))
```

\newpage

# Clustering genes to reduce dimensionality

## Applying k-means classification

```{r}
#GeneX_t <- as.data.frame(t(GeneX))
#GeneX_t_scaled <- as.data.frame(scale(GeneX_t))
GeneX_scaled <- scale(GeneX)
```

```{r}
set.seed(123)
km_genes <- kmeans(GeneX_scaled, centers = 5)

```

## Applying PAM50 classification

```{r}
require(genefu)
require(dplyr)
require(pheatmap)

data(pam50)
data(pam50.robust)

resultat_pam50 <- molecular.subtyping(
  sbt.model = "pam50",
  data = GeneX,
  annot = NULL, 
  do.mapping = FALSE 
)

pam50_df <- as.data.frame(resultat_pam50)

GeneX$subtype <- pam50_df$subtype

head(GeneX$subtype)
```

## Comparing k-means and PAM50 classifications

```{r}
gene_clusters <- data.frame(
  Gene = rownames(GeneX),
  Cluster = km_genes$cluster,
  subtype = pam50_df$subtype
)
```

```{r}
contingency_table <- table(
  "PAM50 Subtype" = gene_clusters$subtype,
  "k-means Cluster" = gene_clusters$Cluster 
  
)

print(contingency_table)
```

Très bons résultats sur tous les subtypes sauf LumA et LumB : c'est normal car il s'agit de types proches ; la classfication k-means fait l'erreur.

| K-means cluster | PAM50 subtype | Precision (with PAM50 as reference) |
|-----------------|---------------|-------------------------------------|
| 1               | Normal        | 0.8456                              |
| 2               | LumB          | 0.6274                              |
| 3               | LumA          | 0.4956                              |
| 4               | Basal         | 0.9171                              |

## Join the GeneX and Clinical_data datasets

```{r}
df <- clinical_data
df$subtype <- pam50_df$subtype
```

## Cleaning the joined dataset

```{r}
# Function to clean dataset (hypothesis : the format is the same as before)
clean_dataset <- function(df) {

  # 1. Suppression des lignes masculines et de la colonne gender
  df <- subset(df, gender != "male")

  # 2. Création et encodage des variables binaires (0/1)
  df$is_dead <- ifelse(df$vital_status == "Dead", 1, 0)
  df$had_prior_treatment <- ifelse(df$prior_treatment == "Yes", 1, 0)
  df$diagnosis_is_primary_disease <- ifelse(df$diagnosis_is_primary_disease == "TRUE", 1, 0)

  # 3. Traitement spécifique de la colonne sites_of_involvement
  df$sites_of_involvement <- sapply(df$sites_of_involvement, function(x) {
    if (is.null(x)) return(NA)      # handle NULLs
    paste(x, collapse = "; ")       # combine multiple strings per row
  })

  # 4. Conversion des colonnes en facteurs
  cols_to_factor <- c(
    "tissue_type",
    "laterality",
    "tissue_or_organ_of_origin",
    "primary_diagnosis",
    "ajcc_pathologic_t",
    "morphology",
    "classification_of_tumor",
    "follow_ups_disease_response",
    "race",
    "sites_of_involvement"
  )
  
  df[cols_to_factor] <- lapply(df[cols_to_factor], factor)
  
  # 5. Suppression des colonnes 
  
  cols_to_remove <- c(
    "gender", #bc all female
    "disease_type", #bc all patients have the same disease_type
    "vital_status", #bc it has been replaced by is_dead
    "prior_treatment",#bc it is now binaery had_prior_tratment
    "patient_id", #bc we don't need it anymore (already joined datasets)
    "age_is_obfuscated", #useless var
    "primary_site", #bc its all Breast
    "days_to_birth", #bc gives NA as coeffs
    "diagnosis_is_primary_disease", #bc gives NA as coeffs
    "classification_of_tumor", #bc gives NA as coeffs
    "bcr_patient_barcode", #useless var
    "ethnicity", #bc causes perfect separation
    "age_at_diagnosis"
  )

  cols_to_keep <- setdiff(names(df), cols_to_remove)
  
  df <- df[, cols_to_keep]
  
  df <- na.omit(df)
}
```

```{r}
df <- clean_dataset(df)
```

```{r}
# Function to drop unused levels in a df
drop_unused_levels_all <- function(df) {
  df_out <- df
  factor_cols <- names(df_out)[sapply(df_out, is.factor)]
  
  for (col in factor_cols) {
    df_out[[col]] <- droplevels(df_out[[col]])
  }
  
  return(df_out)
}
```

```{r}
table(df$tissue_or_organ_of_origin)
```

```{r}
pairs(df[, 1:5])
pairs(df[, 6:10])
pairs(df[, 11:15])
```

```{r}
# Morphology is not useful ; 
# Almost every tissue or organ of origin is "Breast, NOS"
# Follow_ups_disease_response is data_leakage and too deterministic : breaks models
df <- subset(df, select=-c(morphology, 
                           tissue_or_organ_of_origin, 
                           follow_ups_disease_response))

# Only one observation is american indian or alaska native
df <- df[df$race != "american indian or alaska native", ]

require(forcats)

df$primary_diagnosis <- fct_lump_min(df$primary_diagnosis, min=15)
df$ajcc_pathologic_t <- fct_lump_min(df$ajcc_pathologic_t, min=20)
df$sites_of_involvement <- fct_lump_min(df$sites_of_involvement, min=15)

# keep only non-empty levels
df <- drop_unused_levels_all(df)
```

```{r}
# Diagnostic de la répartition de l'outcome binaire par niveaux de facteurs
diagnose_factor_separation <- function(df, outcome_col_name = "is_dead") {
  # Identifier les colonnes de type facteur (y compris les facteurs ordonnés)
  factor_cols <- names(df)[sapply(df, is.factor)]
  
  if (length(factor_cols) == 0) {
    cat("Aucune variable de type facteur n'a été trouvée dans le dataframe.\n")
    return(invisible(NULL))
  }
  
  cat("--- Diagnostic de Séparation par Facteur ---\n")
  cat(paste("Outcome analysé : ", outcome_col_name, "\n\n"))
  
  # 2. Boucle à travers chaque colonne facteur
  for (col_name in factor_cols) {
    
    # Exclure la colonne d'outcome elle-même de la boucle
    if (col_name == outcome_col_name) {
      next
    }
    
    cat(paste("### Facteur :", col_name, "###\n"))
    
    # Créer la table de contingence (Outcome par niveaux de Facteur)
    contingency_table <- table(df[[col_name]], df[[outcome_col_name]], useNA = "ifany")
    
    # Afficher la table
    print(contingency_table)
    
    # 3. Identifier les cas de Séparation Complète
    # Un cas de séparation complète est détecté si une ligne contient un '0' 
    # dans une colonne et un nombre > 0 dans l'autre (i.e., tous les résultats sont du même type).
    
    # Détecter si un niveau n'a QUE des 0 (colonne 1 = 0 et colonne 2 > 0)
    separation_level_0 <- rownames(contingency_table)[contingency_table[, 1] == 0 & contingency_table[, 2] > 0]
    
    # Détecter si un niveau n'a QUE des 1 (colonne 2 = 0 et colonne 1 > 0)
    separation_level_1 <- rownames(contingency_table)[contingency_table[, 2] == 0 & contingency_table[, 1] > 0]
    
    
    if (length(separation_level_0) > 0 || length(separation_level_1) > 0) {
      cat("\n⚠️ ALERTE SÉPARATION COMPLÈTE (GLM NON-RÉGULARISÉ ÉCHOUERA) ⚠️\n")
      if (length(separation_level_0) > 0) {
        cat(paste("-> Niveaux qui prédisent 'Outcome = 1' (count=0 dans Outcome=0): ", 
                  paste(separation_level_0, collapse = ", "), "\n"))
      }
      if (length(separation_level_1) > 0) {
        cat(paste("-> Niveaux qui prédisent 'Outcome = 0' (count=0 dans Outcome=1): ", 
                  paste(separation_level_1, collapse = ", "), "\n"))
      }
    }
    
    cat("\n---\n\n")
  }
  
  return(invisible(NULL))
}
```

```{r}
diagnose_factor_separation(df)
```

```{r}
require(dplyr)
df$id <- rownames(df)
GeneX$id <- rownames(GeneX)

df.complete <- left_join(df, GeneX, by="id")

df <- subset(df, select=-id)
GeneX <- subset(GeneX, select=-id)
```

```{r}
df.complete <- subset(df.complete, select=-c(id, subtype.x, subtype.y))
```

```{r}
require(ggplot2)

ggplot(df, aes(x = factor(is_dead))) +
  geom_bar(fill = c("blue", "red")) +
  labs(title = "Distribution of Vital status",
       x = "Value",
       y = "Count")
```

# Prediction of the survival outcome

## Full regression

```{r}
full_reg <- glm(is_dead ~., data=df, family="binomial")
summary(full_reg)
```

## Regressions avec sélection de variables

```{r}
bw_aic_reg = step(full_reg, direction="backward")
cat("\n\nHere AIC is BIC !\n\n")
bw_bic_reg = step(full_reg, direction="backward", k=log(nrow(df)))
```

```{r}
fw_aic_reg <- step(full_reg, direction="forward")
cat("\n\nHere AIC is BIC !\n\n")
fw_bic_reg <- step(full_reg, direction="forward", k=log(nrow(df)))
```

```{r}
st_aic_reg <- step(full_reg, direction="both")
cat("\n\nHere AIC is BIC !\n\n")
st_bic_reg <- step(full_reg, direction="both", k=log(nrow(df)))
```

## Régression avec pénalisation (Ridge & Lasso)

```{r}
get_performances <- function(y_true, y_probs, threshold = 0.5) {
  
  # Convert probabilities to classes
  y_pred <- ifelse(y_probs >= threshold, 1, 0)
  
  # Confusion matrix
  cm <- table(
    Actual = y_true,
    Predicted = y_pred
  )
  
  # Extract values safely
  TN <- cm["0", "0"]
  FP <- cm["0", "1"]
  FN <- cm["1", "0"]
  TP <- cm["1", "1"]
  
  # Metrics
  accuracy  <- (TP + TN) / sum(cm)
  recall    <- TP / (TP + FN)
  precision <- TP / (TP + FP)
  F_score  <- 2 * (precision * recall) / (precision + recall)
  
  return(list(
    confusion_matrix = cm,
    accuracy = accuracy,
    recall = recall,
    precision = precision,
    F_score = F_score
  ))
}
```

```{r}
optimize_threshold <- function(y_true, y_prob) {
  thresholds <- seq(0.05, 0.6, by = 0.01)
  
  best_F <- -Inf
  best_t <- 0.5
  
  for (t in thresholds) {
    pred <- ifelse(y_prob > t, 1, 0)
    cm <- table(pred, y_true)
    
    if (all(dim(cm) == c(2,2))) {
      TP <- cm[2,2]
      FP <- cm[2,1]
      FN <- cm[1,2]
      
      precision <- TP / (TP + FP)
      recall    <- TP / (TP + FN)
      F1 <- 2 * precision * recall / (precision + recall)
      
      if (!is.nan(F1) && F1 > best_F) {
        best_F <- F1
        best_t <- t
      }
    }
  }
  
  list(best_threshold = best_t, best_F1 = best_F)
}
```

## Simple train / test split

```{r}
require(caret)
set.seed(123)

df$is_dead <- factor(df$is_dead, levels = c(0, 1), labels = c("No", "Yes"))
df.complete$is_dead <- factor(df.complete$is_dead, levels = c(0, 1), labels = c("No", "Yes"))

train_idx <- createDataPartition(df$is_dead, p = 0.8, list = FALSE)
train_data <- df[train_idx, ]
test_data  <- df[-train_idx, ]

train_idx.genes <- createDataPartition(df.complete$is_dead, p = 0.8, list = FALSE)
train_data.genes <- df.complete[train_idx.genes, ]
test_data.genes  <- df.complete[-train_idx.genes, ]
```

```{r}
# -------------------------------
# 1. Outcome vectors (y)
# -------------------------------

# Make sure outcome is numeric 0/1 for glmnet
y_train <- as.numeric(train_data$is_dead == "Yes")
y_test  <- as.numeric(test_data$is_dead  == "Yes")

y_train.genes <- as.numeric(train_data.genes$is_dead == "Yes")
y_test.genes  <- as.numeric(test_data.genes$is_dead  == "Yes")

# -------------------------------
# 2. Predictor data frames (x)
# -------------------------------

x_train <- train_data[, setdiff(names(train_data), "is_dead"), drop = FALSE]
x_test  <- test_data[,  setdiff(names(test_data),  "is_dead"), drop = FALSE]

x_train.genes <- train_data.genes[, setdiff(names(train_data.genes), "is_dead"), drop = FALSE]
x_test.genes  <- test_data.genes[,  setdiff(names(test_data.genes),  "is_dead"), drop = FALSE]

# -------------------------------
# 3. Convert to model matrices (dummy encoding)
# -------------------------------
x_train_mat <- model.matrix(~ ., data = x_train)[, -1]  # remove intercept
x_test_mat  <- model.matrix(~ ., data = x_test)[, -1]

x_train_mat.genes <- model.matrix(~ ., data = x_train.genes)[, -1]
x_test_mat.genes  <- model.matrix(~ ., data = x_test.genes)[, -1]
# -------------------------------
# 4. Force glmnet-compatible format
# -------------------------------

x_train_mat <- as.matrix(x_train_mat)
x_test_mat  <- as.matrix(x_test_mat)

x_train_mat.genes <- as.matrix(x_train_mat.genes)
x_test_mat.genes <- as.matrix(x_test_mat.genes)

# -------------------------------
# 5. Safety check (this MUST be TRUE)
# -------------------------------

stopifnot(ncol(x_train_mat) == ncol(x_test_mat))
```

### Model Weights

```{r}
class_counts <- table(train_data$is_dead)
N <- sum(class_counts)
K <- length(class_counts)

class_weights <- N / (K * class_counts)

model_weights <- ifelse(train_data$is_dead == "Yes",
                        class_weights["Yes"],
                        class_weights["No"])
```

```{r}
# Classic Full Model
full <- glm(
  is_dead ~ ., 
  data = train_data, 
  family = "binomial"
)

full.probs <- predict(full, newdata = test_data, type = "response")

full.opt <- optimize_threshold(y_test, full.probs)

full.perf <- get_performances(y_test, full.probs, full.opt)
```

```{r}
# Full Model weighted
full_weighted <- glm(
  is_dead ~ ., 
  data = train_data, 
  family = "binomial",
  weights = model_weights
)

full_w.probs <- predict(full_weighted, newdata = test_data, type = "response")

full_w.opt <- optimize_threshold(y_test, full_w.probs)

full_w.perf <- get_performances(y_test, full_w.probs, full_w.opt)
```

### Resampling : Combination

```{r}
require(caret)
# balance the train data
train_data.up <- upSample(train_data[,-which(names(train_data)=="is_dead")],
                           train_data$is_dead,
                           yname = "is_dead")

table(train_data.up$is_dead)
table(test_data$is_dead)

train_data.genes.up <- upSample(train_data.genes[,-which(names(train_data.genes)=="is_dead")],
                           train_data.genes$is_dead,
                           yname = "is_dead")

table(train_data.genes.up$is_dead)
table(test_data.genes$is_dead)

test_data.genes.up <- test_data.genes
test_data.up <- test_data
```

```{r}
# -------------------------------
# 1. Outcome vectors (y)
# -------------------------------

# Make sure outcome is numeric 0/1 for glmnet
y_train.up <- as.numeric(train_data.up$is_dead == "Yes")
y_test  <- as.numeric(test_data$is_dead  == "Yes")

y_train.genes.up <- as.numeric(train_data.genes.up$is_dead == "Yes")

# -------------------------------
# 2. Predictor data frames (x)
# -------------------------------

x_train.genes.up <- train_data.genes.up[, setdiff(names(train_data.genes.up), "is_dead"), drop = FALSE]
x_test.genes.up <- test_data.genes.up[, setdiff(names(test_data.genes.up), "is_dead"), drop = FALSE]

x_train.up <- train_data.up[, setdiff(names(train_data.up), "is_dead"), drop = FALSE]
x_test.up <- test_data.up[, setdiff(names(test_data.up), "is_dead"), drop = FALSE]

# -------------------------------
# 3. Convert to model matrices (dummy encoding)
# -------------------------------
x_train_mat.genes.up <- model.matrix(~ ., data = x_train.genes.up)[, -1]
x_test_mat.genes.up  <- model.matrix(~ ., data = x_test.genes.up)[, -1]

x_train_mat.up <- model.matrix(~ ., data = x_train.up)[, -1]
x_test_mat.up  <- model.matrix(~ ., data = x_test.up)[, -1]
# -------------------------------
# 4. Force glmnet-compatible format
# -------------------------------

x_train_mat.genes.up <- as.matrix(x_train_mat.genes.up)
x_test_mat.genes.up <- as.matrix(x_test_mat.genes.up)

x_train_mat.up <- as.matrix(x_train_mat.up)
x_test_mat.up <- as.matrix(x_test_mat.up)
```

```{r}
# Full Model with upsampling
full_up <- glm(
  is_dead ~ ., 
  data = train_data.up, 
  family = "binomial"
)

full_up.probs <- predict(full_up, newdata = test_data, type = "response")

full_up.opt <- optimize_threshold(y_test, full_up.probs)

full_up.perf <- get_performances(y_test, full_up.probs, full_up.opt)
```

### Simple Models

```{r include=FALSE}
full_model <- glm(
  is_dead ~ ., 
  data = train_data, 
  family = "binomial"
)

full_model_probs <- predict(full_model, newdata = test_data, type = "response")

# step() keeps the weights of the fuyll_model so ok
backward_aic_model <- step(full_model, direction = "backward")
backward_aic_probs <- predict(backward_aic_model, newdata = test_data, type = "response")

backward_bic_model <- step(full_model, direction = "backward", k = log(nrow(train_data)))
backward_bic_probs <- predict(backward_bic_model, newdata = test_data, type = "response")

forward_aic_model <- step(full_model, direction = "forward")
forward_aic_probs <- predict(forward_aic_model, newdata = test_data, type = "response")

forward_bic_model <- step(full_model, direction = "forward", k = log(nrow(train_data)))
forward_bic_probs <- predict(forward_bic_model, newdata = test_data, type = "response")

stepwise_aic_model <- step(full_model, direction = "both")
stepwise_aic_probs <- predict(stepwise_aic_model, newdata = test_data, type = "response")

stepwise_bic_model <- step(full_model, direction = "both", k = log(nrow(train_data)))
stepwise_bic_probs <- predict(stepwise_bic_model, newdata = test_data, type = "response")
```

```{r}

```

### Advanced Models

```{r}
class_counts.genes <- table(train_data.genes$is_dead)
N.genes <- sum(class_counts.genes)
K.genes <- length(class_counts.genes)

class_weights.genes <- N.genes / (K.genes * class_counts.genes)

model_weights.genes <- ifelse(train_data.genes$is_dead == "Yes",
                        class_weights.genes["Yes"],
                        class_weights.genes["No"])
```

## Ridge Regression Models (alpha=0)

### BASE RIDGE - WITH CANCER SUBTYPE

# RIDGE: Clinical + Subtype Data (K-clustered)

```{r}
require(glmnet)
ridge_reg <- glmnet(x_train_mat, y_train, alpha = 0, family="binomial", standardize=TRUE)

ridge_cv_model <- cv.glmnet(x_train_mat, y_train, alpha = 0, n_folds=10, family="binomial", standardize = TRUE)

ridge_min.lambda <- ridge_cv_model$lambda.min
ridge_1se.lambda <- ridge_cv_model$lambda.1se 
cat("Lambda that minimizes RSE (Ridge K) : ", ridge_min.lambda, "\nLambda 1SE (Ridge K) : ", ridge_1se.lambda)
```

```{r}
ridge_min <- glmnet(x_train_mat, y_train, alpha = 0, lambda = ridge_min.lambda, family="binomial",standardize = TRUE)
ridge_1se <- glmnet(x_train_mat, y_train, alpha = 0, lambda = ridge_1se.lambda, family="binomial", standardize = TRUE)
```

```{r}
ridge_min.probs <- predict(ridge_min, newx = x_test_mat, type = "response")

ridge_min.opt_treshold <- optimize_threshold(y_test, ridge_min.probs)

ridge_min.perf <- get_performances( y_true = y_test, y_probs = ridge_min.probs, threshold = ridge_min.opt_treshold$best_threshold )

ridge_1se.probs <- predict(ridge_1se, newx = x_test_mat, type = "response")

ridge_1se.opt_treshold <- optimize_threshold(y_test, ridge_1se.probs)

ridge_1se.perf <- get_performances( y_true = y_test, y_probs = ridge_1se.probs, threshold = ridge_1se.opt_treshold$best_threshold )
```

### BASE RIDGE - WITH ALL GENES

```{r}
# RIDGE: Clinical + All Genes Data (C-complete)

ridge_reg.genes <- glmnet(x_train_mat.genes, y_train.genes, alpha = 0, family="binomial", standardize=TRUE)

ridge_cv_model.genes <- cv.glmnet(x_train_mat.genes, y_train.genes, alpha = 0, n_folds=10, family="binomial", standardize = TRUE)

ridge_min_lambda.genes <- ridge_cv_model.genes$lambda.min
ridge_1se_lambda.genes <- ridge_cv_model.genes$lambda.1se
cat("Lambda that minimizes RSE (Ridge C) : ", ridge_min_lambda.genes, "\nLambda 1SE (Ridge C) : ", ridge_1se_lambda.genes)
```

```{r}
ridge_min.genes <- glmnet(x_train_mat.genes, y_train.genes, alpha = 0, lambda = ridge_min_lambda.genes, family="binomial",standardize = TRUE)
ridge_1se.genes <- glmnet(x_train_mat.genes, y_train.genes, alpha = 0, lambda = ridge_1se_lambda.genes, family="binomial", standardize = TRUE)
```

```{r}
ridge_min.genes.probs <- predict(ridge_min.genes, newx = x_test_mat.genes, type = "response")

ridge_min.genes.opt_treshold <- optimize_threshold(y_test.genes, ridge_min.genes.probs)

ridge_min.genes.perf <- get_performances( y_true = y_test.genes, y_probs = ridge_min.genes.probs, threshold = ridge_min.genes.opt_treshold$best_threshold )

ridge_1se.genes.probs <- predict(ridge_1se.genes, newx = x_test_mat.genes, type = "response")

ridge_1se.genes.opt_treshold <- optimize_threshold(y_test.genes, ridge_1se.genes.probs)

ridge_1se.genes.perf <- get_performances( y_true = y_test.genes, y_probs = ridge_1se.genes.probs, threshold = ridge_1se.genes.opt_treshold$best_threshold )
```

### RIDGE UPSAMPLED - WITH ALL GENES

```{r}
# RIDGE: Clinical + All Genes Data (C-complete) Upsampled

ridge_reg.genes.up <- glmnet(x_train_mat.genes.up, y_train.genes.up, alpha = 0, family="binomial", standardize=TRUE)

ridge_cv_model.genes.up <- cv.glmnet(x_train_mat.genes.up, y_train.genes.up, alpha = 0, n_folds=10, family="binomial", standardize = TRUE)

ridge_min_lambda.genes.up <- ridge_cv_model.genes.up$lambda.min
ridge_1se_lambda.genes.up <- ridge_cv_model.genes.up$lambda.1se 
cat("Lambda that minimizes RSE (Ridge C Up) : ", ridge_min_lambda.genes.up, "\nLambda 1SE (Ridge C Up) : ", ridge_1se_lambda.genes.up)
```

```{r}
ridge_min.genes.up <- glmnet(x_train_mat.genes.up, y_train.genes.up, alpha = 0, lambda = ridge_min_lambda.genes.up, family="binomial",standardize = TRUE) 
ridge_1se.genes.up <- glmnet(x_train_mat.genes.up, y_train.genes.up, alpha = 0, lambda = ridge_1se_lambda.genes.up, family="binomial", standardize = TRUE)
```

```{r}
ridge_min.genes.up.probs <- predict(ridge_min.genes.up, newx = x_test_mat.genes.up, type = "response")

ridge_min.genes.up.opt_treshold <- optimize_threshold(y_test.genes, ridge_min.genes.up.probs)

ridge_min.genes.up.perf <- get_performances( y_true = y_test.genes, y_probs = ridge_min.genes.up.probs, threshold = ridge_min.genes.up.opt_treshold$best_threshold )

ridge_1se.genes.up.probs <- predict(ridge_1se.genes.up, newx = x_test_mat.genes.up, type = "response")

ridge_1se.genes.up.opt_treshold <- optimize_threshold(y_test.genes, ridge_1se.genes.up.probs)

ridge_1se.genes.up.perf <- get_performances( y_true = y_test.genes, y_probs = ridge_1se.genes.up.probs, threshold = ridge_1se.genes.up.opt_treshold$best_threshold )
```

### RIDGE UPSAMPLED - WITH CANCER SUBTYPE

```{r}
# RIDGE: Clinical + Subtype Data (K-clustered) Upsampled

ridge_reg.up <- glmnet(x_train_mat.up, y_train.up, alpha = 0, family="binomial", standardize=TRUE)

ridge_cv_model.up <- cv.glmnet(x_train_mat.up, y_train.up, alpha = 0, n_folds=10, family="binomial", standardize = TRUE)

ridge_min_lambda.up <- ridge_cv_model.up$lambda.min
ridge_1se_lambda.up <- ridge_cv_model.up$lambda.1se 
cat("Lambda that minimizes RSE (Ridge K Up) : ", ridge_min_lambda.up, "\nLambda 1SE (Ridge K Up) : ", ridge_1se_lambda.up)
```

```{r}
ridge_min.up <- glmnet(x_train_mat.up, y_train.up, alpha = 0, lambda = ridge_min_lambda.up, family="binomial",standardize = TRUE) 
ridge_1se.up <- glmnet(x_train_mat.up, y_train.up, alpha = 0, lambda = ridge_1se_lambda.up, family="binomial", standardize = TRUE)
```

```{r}
ridge_min.up.probs <- predict(ridge_min.up, newx = x_test_mat.up, type = "response")

ridge_min.up.opt_treshold <- optimize_threshold(y_test.genes, ridge_min.up.probs)

ridge_min.up.perf <- get_performances( y_true = y_test.genes, y_probs = ridge_min.up.probs, threshold = ridge_min.up.opt_treshold$best_threshold )

ridge_1se.up.probs <- predict(ridge_1se.up, newx = x_test_mat.up, type = "response")

ridge_1se.up.opt_treshold <- optimize_threshold(y_test.genes, ridge_1se.up.probs)

ridge_1se.up.perf <- get_performances( y_true = y_test.genes, y_probs = ridge_1se.up.probs, threshold = ridge_1se.up.opt_treshold$best_threshold )
```

### BASE RIDGE WITH WEIGHTS (K-clustered)

```{r}
# RIDGE: Clinical + Subtype Data (K-clustered) Weighted

ridge_reg.weighted <- glmnet(x_train_mat, y_train, alpha = 0, family="binomial", standardize=TRUE, weights = model_weights)

ridge_cv_model.weighted <- cv.glmnet(x_train_mat, y_train, alpha = 0, n_folds=10, family="binomial", standardize = TRUE, weights = model_weights)
```

```{r}
ridge_min_lambda.weighted <- ridge_cv_model.weighted$lambda.min
ridge_1se_lambda.weighted <- ridge_cv_model.weighted$lambda.1se 
cat("Lambda that minimizes RSE (Ridge K W) : ", ridge_min_lambda.weighted, "\nLambda 1SE (Ridge K W) : ", ridge_1se_lambda.weighted)

ridge_min.weighted <- glmnet(x_train_mat, y_train, alpha = 0, lambda = ridge_min_lambda.weighted, family="binomial", standardize = TRUE, weights = model_weights)
ridge_1se.weighted <- glmnet(x_train_mat, y_train, alpha = 0, lambda = ridge_1se_lambda.weighted, family="binomial", standardize = TRUE, weights = model_weights)
```

```{r}
ridge_min.weighted.probs <- predict(ridge_min.weighted, newx = x_test_mat, type = "response")

ridge_min.weighted.opt_treshold <- optimize_threshold(y_test, ridge_min.weighted.probs)

ridge_min.weighted.perf <- get_performances( y_true = y_test, y_probs = ridge_min.weighted.probs, threshold = ridge_min.weighted.opt_treshold$best_threshold )

ridge_1se.weighted.probs <- predict(ridge_1se.weighted, newx = x_test_mat, type = "response")

ridge_1se.weighted.opt_treshold <- optimize_threshold(y_test, ridge_1se.weighted.probs)

ridge_1se.weighted.perf <- get_performances( y_true = y_test, y_probs = ridge_1se.weighted.probs, threshold = ridge_1se.weighted.opt_treshold$best_threshold )
```

### RIDGE WITH WEIGHTS - WITH ALL GENES (C-complete)

```{r}
# RIDGE: Clinical + All Genes Data (C-complete) Weighted

ridge_reg.genes.weighted <- glmnet(x_train_mat.genes, y_train.genes, alpha = 0, family="binomial", standardize=TRUE, weights = model_weights.genes)

ridge_cv_model.genes.weighted <- cv.glmnet(x_train_mat.genes, y_train.genes, alpha = 0, n_folds=10, family="binomial", standardize = TRUE, weights = model_weights.genes)
```

```{r}
ridge_min_lambda.genes.weighted <- ridge_cv_model.genes.weighted$lambda.min
ridge_1se_lambda.genes.weighted <- ridge_cv_model.genes.weighted$lambda.1se 
cat("Lambda that minimizes RSE (Ridge C W) : ", ridge_min_lambda.genes.weighted, "\nLambda 1SE (Ridge C W) : ", ridge_1se_lambda.genes.weighted)
```

```{r}
ridge_min.genes.weighted <- glmnet(x_train_mat.genes, y_train.genes, alpha = 0, lambda = ridge_min_lambda.genes.weighted, family="binomial", standardize = TRUE, weights = model_weights.genes)
ridge_1se.genes.weighted <- glmnet(x_train_mat.genes, y_train.genes, alpha = 0, lambda = ridge_1se_lambda.genes.weighted, family="binomial", standardize = TRUE, weights = model_weights.genes)
```

```{r}
ridge_min.genes.weighted.probs <- predict(ridge_min.genes.weighted, newx = x_test_mat.genes, type = "response")

ridge_min.genes.weighted.opt_treshold <- optimize_threshold(y_test.genes, ridge_min.genes.weighted.probs)

ridge_min.genes.weighted.perf <- get_performances( y_true = y_test.genes, y_probs = ridge_min.genes.weighted.probs, threshold = ridge_min.genes.weighted.opt_treshold$best_threshold )

ridge_1se.genes.weighted.probs <- predict(ridge_1se.genes.weighted, newx = x_test_mat.genes, type = "response")

ridge_1se.genes.weighted.opt_treshold <- optimize_threshold(y_test.genes, ridge_1se.genes.weighted.probs)

ridge_1se.genes.weighted.perf <- get_performances( y_true = y_test.genes, y_probs = ridge_1se.genes.weighted.probs, threshold = ridge_1se.genes.weighted.opt_treshold$best_threshold )
```

### BASE LASSO - WITH CANCER SUBTYPE

```{r}
lasso_reg <- glmnet(x_train_mat, y_train, alpha = 1, 
                    family="binomial", standardize=TRUE)

lasso_cv_model <- cv.glmnet(x_train_mat, y_train, alpha = 1, 
                            n_folds=10, family="binomial",
                            standardize = TRUE)

lasso_min.lambda <- lasso_cv_model$lambda.min
lasso_1se.lambda <- lasso_cv_model$lambda.1se
cat("Lambda that minimizes RSE : ", lasso_min.lambda, "\nLambda 1SE : ", lasso_1se.lambda)
```

```{r}
lasso_min <- glmnet(x_train_mat, y_train, alpha = 1, lambda = lasso_min.lambda, 
                    family="binomial",standardize = TRUE)
lasso_1se <- glmnet(x_train_mat, y_train, alpha = 1, lambda = lasso_1se.lambda, 
                    family="binomial", standardize = TRUE)
```

```{r}
lasso_min.probs <- predict(lasso_min, newx = x_test_mat, type = "response")

lasso_min.opt_treshold <- optimize_threshold(y_test, lasso_min.probs)

lasso_min.perf <- get_performances(
  y_true  = y_test,
  y_probs = lasso_min.probs,
  threshold = lasso_min.opt_treshold$best_threshold
)

lasso_1se.probs <- predict(lasso_1se, newx = x_test_mat, type = "response")

lasso_1se.opt_treshold <- optimize_threshold(y_test, lasso_1se.probs)

lasso_1se.perf <- get_performances(
  y_true  = y_test,
  y_probs = lasso_1se.probs,
  threshold = lasso_1se.opt_treshold$best_threshold
)
```

### BASE LASSO - WITH ALL GENES

```{r}
# TRY WITH THE COMPLETE DATSET (USE GENES EXPRESSION)
lasso_reg.genes <- glmnet(x_train_mat.genes, y_train.genes, alpha = 1, 
                    family="binomial", standardize=TRUE)

lasso_cv_model.genes <- cv.glmnet(x_train_mat.genes, y_train.genes, alpha = 1, 
                            n_folds=10, family="binomial",
                            standardize = TRUE)

lasso_min_lambda.genes <- lasso_cv_model.genes$lambda.min
lasso_1se_lambda.genes <- lasso_cv_model.genes$lambda.1se
cat("Lambda that minimizes RSE : ", lasso_min_lambda.genes, "\nLambda 1SE : ", lasso_1se_lambda.genes)
```

```{r}
lasso_min.genes <- glmnet(x_train_mat.genes, y_train.genes, 
                          alpha = 1, lambda = lasso_min_lambda.genes, 
                          family="binomial",standardize = TRUE)
lasso_1se.genes <- glmnet(x_train_mat.genes, y_train.genes, 
                          alpha = 1, lambda = lasso_1se_lambda.genes, 
                          family="binomial", standardize = TRUE)
```

```{r}
lasso_min.genes.probs <- predict(lasso_min.genes, newx = x_test_mat.genes, type = "response")

lasso_min.genes.opt_treshold <- optimize_threshold(y_test.genes, lasso_min.genes.probs)

lasso_min.genes.perf <- get_performances(
  y_true  = y_test.genes,
  y_probs = lasso_min.genes.probs,
  threshold = lasso_min.genes.opt_treshold$best_threshold
)

lasso_1se.genes.probs <- predict(lasso_1se.genes, newx = x_test_mat.genes, type = "response")

lasso_1se.genes.opt_treshold <- optimize_threshold(y_test.genes, lasso_1se.genes.probs)

lasso_1se.genes.perf <- get_performances(
  y_true  = y_test.genes,
  y_probs = lasso_1se.genes.probs,
  threshold = lasso_1se.genes.opt_treshold$best_threshold
)
```

### LASSO UPSAMPLED - WITH ALL GENES

```{r}
# TRY WITH THE COMPLETE DATASET UPSAMPLED (USE GENES EXPRESSION)
lasso_reg.genes.up <- glmnet(x_train_mat.genes.up, y_train.genes.up, alpha = 1, 
                    family="binomial", standardize=TRUE)

lasso_cv_model.genes.up <- cv.glmnet(x_train_mat.genes.up, y_train.genes.up, alpha = 1, 
                            n_folds=10, family="binomial",
                            standardize = TRUE)

lasso_min_lambda.genes.up <- lasso_cv_model.genes.up$lambda.min
lasso_1se_lambda.genes.up <- lasso_cv_model.genes.up$lambda.1se
cat("Lambda that minimizes RSE : ", lasso_min_lambda.genes.up, "\nLambda 1SE : ", lasso_1se_lambda.genes.up)
```

```{r}
lasso_min.genes.up <- glmnet(x_train_mat.genes.up, y_train.genes.up, 
                          alpha = 1, lambda = lasso_min_lambda.genes.up, 
                          family="binomial",standardize = TRUE)
lasso_1se.genes.up <- glmnet(x_train_mat.genes.up, y_train.genes.up, 
                          alpha = 1, lambda = lasso_1se_lambda.genes.up, 
                          family="binomial", standardize = TRUE)
```

```{r}
lasso_min.genes.up.probs <- predict(lasso_min.genes.up, newx = x_test_mat.genes.up, type = "response")

lasso_min.genes.up.opt_treshold <- optimize_threshold(y_test.genes, lasso_min.genes.up.probs)

lasso_min.genes.up.perf <- get_performances(
  y_true  = y_test.genes,
  y_probs = lasso_min.genes.up.probs,
  threshold = lasso_min.genes.up.opt_treshold$best_threshold
)

lasso_1se.genes.up.probs <- predict(lasso_1se.genes.up, newx = x_test_mat.genes.up, type = "response")

lasso_1se.genes.up.opt_treshold <- optimize_threshold(y_test.genes, lasso_1se.genes.up.probs)

lasso_1se.genes.up.perf <- get_performances(
  y_true  = y_test.genes,
  y_probs = lasso_1se.genes.up.probs,
  threshold = lasso_1se.genes.up.opt_treshold$best_threshold
)
```

### LASSO UPSAMPLED - WITH CANCER SUBTYPE

```{r}
# TRY WITH THE KMEANS DATASET UPSAMPLED
lasso_reg.up <- glmnet(x_train_mat.up, y_train.up, alpha = 1, 
                    family="binomial", standardize=TRUE)

lasso_cv_model.up <- cv.glmnet(x_train_mat.up, y_train.up, alpha = 1, 
                            n_folds=10, family="binomial",
                            standardize = TRUE)

lasso_min_lambda.up <- lasso_cv_model.up$lambda.min
lasso_1se_lambda.up <- lasso_cv_model.up$lambda.1se
cat("Lambda that minimizes RSE : ", lasso_min_lambda.up, "\nLambda 1SE : ", lasso_1se_lambda.up)
```

```{r}
lasso_min.up <- glmnet(x_train_mat.up, y_train.up, 
                          alpha = 1, lambda = lasso_min_lambda.up, 
                          family="binomial",standardize = TRUE)
lasso_1se.up <- glmnet(x_train_mat.up, y_train.up, 
                          alpha = 1, lambda = lasso_1se_lambda.up, 
                          family="binomial", standardize = TRUE)
```

```{r}
lasso_min.up.probs <- predict(lasso_min.up, newx = x_test_mat.up, type = "response")

lasso_min.up.opt_treshold <- optimize_threshold(y_test.genes, lasso_min.up.probs)

lasso_min.up.perf <- get_performances(
  y_true  = y_test.genes,
  y_probs = lasso_min.up.probs,
  threshold = lasso_min.up.opt_treshold$best_threshold
)

lasso_1se.up.probs <- predict(lasso_1se.up, newx = x_test_mat.up, type = "response")

lasso_1se.up.opt_treshold <- optimize_threshold(y_test.genes, lasso_1se.up.probs)

lasso_1se.up.perf <- get_performances(
  y_true  = y_test.genes,
  y_probs = lasso_1se.up.probs,
  threshold = lasso_1se.up.opt_treshold$best_threshold
)
```

### BASE LASSO WITH WEIGHTS

```{r}
# BASE LASSO WITH WEIGHTS
lasso_reg.weighted <- glmnet(x_train_mat, y_train, alpha = 1, 
                    family="binomial", standardize=TRUE, weights = model_weights)

lasso_cv_model.weighted <- cv.glmnet(x_train_mat, y_train, alpha = 1, 
                            n_folds=10, family="binomial",
                            standardize = TRUE, weights = model_weights)

lasso_min_lambda.weighted <- lasso_cv_model.weighted$lambda.min
lasso_1se_lambda.weighted <- lasso_cv_model.weighted$lambda.1se
cat("Lambda that minimizes RSE : ", lasso_min_lambda.weighted, "\nLambda 1SE : ", lasso_1se_lambda.weighted)
```

```{r}
lasso_min.weighted <- glmnet(x_train_mat, y_train, 
                          alpha = 1, lambda = lasso_min_lambda.weighted, 
                          family="binomial", 
                          standardize = TRUE, weights = model_weights)
lasso_1se.weighted <- glmnet(x_train_mat, y_train, 
                          alpha = 1, lambda = lasso_1se_lambda.weighted, 
                          family="binomial", 
                          standardize = TRUE, weights = model_weights)
```

```{r}
lasso_min.weighted.probs <- predict(lasso_min.weighted, newx = x_test_mat, type = "response")

lasso_min.weighted.opt_treshold <- optimize_threshold(y_test, lasso_min.weighted.probs)

lasso_min.weighted.perf <- get_performances(
  y_true  = y_test,
  y_probs = lasso_min.weighted.probs,
  threshold = lasso_min.weighted.opt_treshold$best_threshold
)

lasso_1se.weighted.probs <- predict(lasso_1se.weighted, newx = x_test_mat, type = "response")

lasso_1se.weighted.opt_treshold <- optimize_threshold(y_test, lasso_1se.weighted.probs)

lasso_1se.weighted.perf <- get_performances(
  y_true  = y_test,
  y_probs = lasso_1se.weighted.probs,
  threshold = lasso_1se.weighted.opt_treshold$best_threshold
)
```

### LASSO WITH WEIGHTS - WITH ALL GENES

```{r}
# LASSO WITH WEIGHTS AND ALL GENES
lasso_reg.genes.weighted <- glmnet(x_train_mat.genes, y_train.genes, alpha = 1, 
                    family="binomial", standardize=TRUE, 
                    weights = model_weights.genes)

lasso_cv_model.genes.weighted <- cv.glmnet(x_train_mat.genes, y_train.genes, 
                            alpha = 1, 
                            n_folds=10, family="binomial",
                            standardize = TRUE, weights = model_weights.genes)

lasso_min_lambda.genes.weighted <- lasso_cv_model.genes.weighted$lambda.min
lasso_1se_lambda.genes.weighted <- lasso_cv_model.genes.weighted$lambda.1se
cat("Lambda that minimizes RSE : ", lasso_min_lambda.genes.weighted, "\nLambda 1SE : ", lasso_1se_lambda.genes.weighted)
```

```{r}
lasso_min.genes.weighted <- glmnet(x_train_mat.genes, y_train.genes, 
                          alpha = 1, lambda = lasso_min_lambda.genes.weighted, 
                          family="binomial", 
                          standardize = TRUE, weights = model_weights.genes)
lasso_1se.genes.weighted <- glmnet(x_train_mat.genes, y_train.genes, 
                          alpha = 1, lambda = lasso_1se_lambda.genes.weighted, 
                          family="binomial", 
                          standardize = TRUE, weights = model_weights.genes)
```

```{r}
lasso_min.genes.weighted.probs <- predict(lasso_min.genes.weighted, 
                                          newx = x_test_mat.genes, type = "response")

lasso_min.genes.weighted.opt_treshold <- optimize_threshold(y_test.genes, lasso_min.genes.weighted.probs)

lasso_min.genes.weighted.perf <- get_performances(
  y_true  = y_test.genes,
  y_probs = lasso_min.genes.weighted.probs,
  threshold = lasso_min.genes.weighted.opt_treshold$best_threshold
)

lasso_1se.genes.weighted.probs <- predict(lasso_1se.genes.weighted, newx = x_test_mat.genes, type = "response")

lasso_1se.genes.weighted.opt_treshold <- optimize_threshold(y_test.genes, lasso_1se.genes.weighted.probs)

lasso_1se.genes.weighted.perf <- get_performances(
  y_true  = y_test.genes,
  y_probs = lasso_1se.genes.weighted.probs,
  threshold = lasso_1se.genes.weighted.opt_treshold$best_threshold
)
```

## Elastic Net Regression Models (Alpha Tuning Per Case)

```{r}
alpha_grid <- seq(0, 1, by = 0.05)
```

```{r}
library(glmnet)
set.seed(42)

find_optimal_alpha_lambda <- function(x, y, alpha_grid, k_folds = 10, weights = NULL) {
  
  foldid <- sample(rep(seq(k_folds), length.out = length(y)))
  
  best_alpha <- NA
  best_lambda <- NA
  min_error <- Inf
  
  for (a in alpha_grid) {
    
    cv_fit <- cv.glmnet(
      x, y,
      alpha = a,
      foldid = foldid,
      family = "binomial",
      type.measure = "deviance",
      standardize = TRUE,
      weights = weights # Handles weighted cases
    )
    
    curr_error <- min(cv_fit$cvm)
    
    if (curr_error < min_error) {
      min_error <- curr_error
      best_alpha <- a
      lambda_min <- cv_fit$lambda.min
      lambda_1se <- cv_fit$lamba.1se
    }
  }
  
  list(best_alpha = best_alpha, lambda_min = lambda_min, lambda_1se = lambda_1se)
}
```

```{r}
enet.params <- find_optimal_alpha_lambda(x_train_mat, y_train, alpha_grid)
enet.up.params <- find_optimal_alpha_lambda(x_train_mat.up, y_train.up, alpha_grid)
enet.weights.params <- find_optimal_alpha_lambda(x_train_mat, y_train, alpha_grid, weights = model_weights)

enet.genes.params <- find_optimal_alpha_lambda(x_train_mat.genes, y_train.genes, alpha_grid)
enet.genes.up.params <- find_optimal_alpha_lambda(x_train_mat.genes.up, y_train.genes.up, alpha_grid)
enet.genes.weights.params <- find_optimal_alpha_lambda(x_train_mat.genes, y_train.genes, alpha_grid, weights = model_weights.genes)
```

### BASE EN - WITH CANCER SUBTYPE

```{r}
enet_reg <- glmnet(
  x_train_mat, y_train,
  alpha = enet.params$best_alpha,
  family = "binomial", standardize = TRUE
)

enet_cv_model <- cv.glmnet(
  x_train_mat, y_train,
  alpha = enet.params$best_alpha,
  n_folds = 10, family = "binomial",
  standardize = TRUE
)

enet_min.lambda <- enet_cv_model$lambda.min
enet_1se.lambda <- enet_cv_model$lambda.1se
cat("Lambda min:", enet_min.lambda, "\nLambda 1SE:", enet_1se.lambda)
```

```{r}
enet_min <- glmnet(
  x_train_mat, y_train,
  alpha = enet.params$best_alpha,
  lambda = enet_min.lambda,
  family="binomial", standardize=TRUE
)

enet_1se <- glmnet(
  x_train_mat, y_train,
  alpha = enet.params$best_alpha,
  lambda = enet_1se.lambda,
  family="binomial", standardize=TRUE
)
```

```{r}
enet_min.probs <- predict(enet_min, newx = x_test_mat, type="response")
enet_min.opt_treshold <- optimize_threshold(y_test, enet_min.probs)

enet_min.perf <- get_performances(
  y_true=y_test,
  y_probs=enet_min.probs,
  threshold=enet_min.opt_treshold$best_threshold
)

enet_1se.probs <- predict(enet_1se, newx = x_test_mat, type="response")
enet_1se.opt_treshold <- optimize_threshold(y_test, enet_1se.probs)

enet_1se.perf <- get_performances(
  y_true=y_test,
  y_probs=enet_1se.probs,
  threshold=enet_1se.opt_treshold$best_threshold
)
```

```{r}
enet_min.probs <- predict(enet_min, newx = x_test_mat, type="response")
enet_min.opt_treshold <- optimize_threshold(y_test, enet_min.probs)

enet_min.perf <- get_performances(
  y_true=y_test,
  y_probs=enet_min.probs,
  threshold=enet_min.opt_treshold$best_threshold
)

enet_1se.probs <- predict(enet_1se, newx = x_test_mat, type="response")
enet_1se.opt_treshold <- optimize_threshold(y_test, enet_1se.probs)

enet_1se.perf <- get_performances(
  y_true=y_test,
  y_probs=enet_1se.probs,
  threshold=enet_1se.opt_treshold$best_threshold
)
```

### BASE EN - WITH ALL GENES

```{r}
enet_reg.genes <- glmnet(
  x_train_mat.genes, y_train.genes,
  alpha = enet.genes.params$best_alpha,
  family="binomial", standardize=TRUE
)

enet_cv_model.genes <- cv.glmnet(
  x_train_mat.genes, y_train.genes,
  alpha = enet.genes.params$best_alpha,
  n_folds=10, family="binomial", standardize=TRUE
)

enet_min_lambda.genes <- enet_cv_model.genes$lambda.min
enet_1se_lambda.genes <- enet_cv_model.genes$lambda.1se
```

```{r}
enet_min.genes <- glmnet(
  x_train_mat.genes, y_train.genes,
  alpha = enet.genes.params$best_alpha,
  lambda = enet_min_lambda.genes,
  family="binomial", standardize=TRUE
)

enet_1se.genes <- glmnet(
  x_train_mat.genes, y_train.genes,
  alpha = enet.genes.params$best_alpha,
  lambda = enet_1se_lambda.genes,
  family="binomial", standardize=TRUE
)
```

```{r}
enet_min.genes.probs <- predict(enet_min.genes, newx=x_test_mat.genes, type="response")
enet_min.genes.opt_treshold <- optimize_threshold(y_test.genes, enet_min.genes.probs)

enet_min.genes.perf <- get_performances(
  y_true = y_test.genes,
  y_probs = enet_min.genes.probs,
  threshold = enet_min.genes.opt_treshold$best_threshold
)

enet_1se.genes.probs <- predict(enet_1se.genes, newx=x_test_mat.genes, type="response")
enet_1se.genes.opt_treshold <- optimize_threshold(y_test.genes, enet_1se.genes.probs)

enet_1se.genes.perf <- get_performances(
  y_true = y_test.genes,
  y_probs = enet_1se.genes.probs,
  threshold = enet_1se.genes.opt_treshold$best_threshold
)
```

### EN UPSAMPLED - WITH ALL GENES

```{r}
enet_reg.genes.up <- glmnet(
  x_train_mat.genes.up, y_train.genes.up,
  alpha = enet.genes.up.params$best_alpha,
  family="binomial", standardize=TRUE
)

enet_cv_model.genes.up <- cv.glmnet(
  x_train_mat.genes.up, y_train.genes.up,
  alpha = enet.genes.up.params$best_alpha,
  n_folds=10, family="binomial", standardize=TRUE
)

enet_min_lambda.genes.up <- enet_cv_model.genes.up$lambda.min
enet_1se_lambda.genes.up <- enet_cv_model.genes.up$lambda.1se
```

```{r}
enet_min.genes.up <- glmnet(
  x_train_mat.genes.up, y_train.genes.up,
  alpha = enet.genes.up.params$best_alpha,
  lambda = enet_min_lambda.genes.up,
  family="binomial", standardize=TRUE
)

enet_1se.genes.up <- glmnet(
  x_train_mat.genes.up, y_train.genes.up,
  alpha = enet.genes.up.params$best_alpha,
  lambda = enet_1se_lambda.genes.up,
  family="binomial", standardize=TRUE
)
```

```{r}
enet_min.genes.up.probs <- predict(enet_min.genes.up, newx=x_test_mat.genes.up, type="response")
enet_min.genes.up.opt_treshold <- optimize_threshold(y_test.genes, enet_min.genes.up.probs)

enet_min.genes.up.perf <- get_performances(
  y_true=y_test.genes,
  y_probs=enet_min.genes.up.probs,
  threshold=enet_min.genes.up.opt_treshold$best_threshold
)

enet_1se.genes.up.probs <- predict(enet_1se.genes.up, newx=x_test_mat.genes.up, type="response")
enet_1se.genes.up.opt_treshold <- optimize_threshold(y_test.genes, enet_1se.genes.up.probs)

enet_1se.genes.up.perf <- get_performances(
  y_true=y_test.genes,
  y_probs=enet_1se.genes.up.probs,
  threshold=enet_1se.genes.up.opt_treshold$best_threshold
)
```

### EN UPSAMPLED - WITH CANCER SUBTYPE

```{r}
enet_reg.up <- glmnet(
  x_train_mat.up, y_train.up,
  alpha = enet.up.params$best_alpha,
  family="binomial", standardize=TRUE
)

enet_cv_model.up <- cv.glmnet(
  x_train_mat.up, y_train.up,
  alpha = enet.up.params$best_alpha,
  n_folds=10, family="binomial", standardize=TRUE
)

enet_min_lambda.up <- enet_cv_model.up$lambda.min
enet_1se_lambda.up <- enet_cv_model.up$lambda.1se
```

```{r}
enet_min.up <- glmnet(
  x_train_mat.up, y_train.up,
  alpha = enet.up.params$best_alpha,
  lambda = enet_min_lambda.up,
  family="binomial", standardize=TRUE
)

enet_1se.up <- glmnet(
  x_train_mat.up, y_train.up,
  alpha = enet.up.params$best_alpha,
  lambda = enet_1se_lambda.up,
  family="binomial", standardize=TRUE
)
```

```{r}
enet_min.up.probs <- predict(enet_min.up, newx=x_test_mat.up, type="response")
enet_min.up.opt_treshold <- optimize_threshold(y_test, enet_min.up.probs)

enet_min.up.perf <- get_performances(
  y_true=y_test,
  y_probs=enet_min.up.probs,
  threshold=enet_min.up.opt_treshold$best_threshold
)

enet_1se.up.probs <- predict(enet_1se.up, newx=x_test_mat.up, type="response")
enet_1se.up.opt_treshold <- optimize_threshold(y_test, enet_1se.up.probs)

enet_1se.up.perf <- get_performances(
  y_true=y_test,
  y_probs=enet_1se.up.probs,
  threshold=enet_1se.up.opt_treshold$best_threshold
)
```

### EN WITH WEIGHTS - WITH CANCER SUBTYPE

```{r}
enet_reg.weighted <- glmnet(
  x_train_mat, y_train,
  alpha = enet.weights.params$best_alpha,
  family="binomial", standardize=TRUE,
  weights = model_weights
)

enet_cv_model.weighted <- cv.glmnet(
  x_train_mat, y_train,
  alpha = enet.weights.params$best_alpha,
  n_folds=10, family="binomial",
  standardize=TRUE,
  weights = model_weights
)

enet_min_lambda.weighted <- enet_cv_model.weighted$lambda.min
enet_1se_lambda.weighted <- enet_cv_model.weighted$lambda.1se
```

```{r}
enet_min.weighted <- glmnet(
  x_train_mat, y_train,
  alpha = enet.weights.params$best_alpha,
  lambda = enet_min_lambda.weighted,
  family="binomial", standardize=TRUE,
  weights = model_weights
)

enet_1se.weighted <- glmnet(
  x_train_mat, y_train,
  alpha = enet.weights.params$best_alpha,
  lambda = enet_1se_lambda.weighted,
  family="binomial", standardize=TRUE,
  weights = model_weights
)
```

```{r}
enet_min.weighted.probs <- predict(enet_min.weighted, newx=x_test_mat, type="response")
enet_min.weighted.opt_treshold <- optimize_threshold(y_test, enet_min.weighted.probs)

enet_min.weighted.perf <- get_performances(
  y_true=y_test,
  y_probs=enet_min.weighted.probs,
  threshold=enet_min.weighted.opt_treshold$best_threshold
)

enet_1se.weighted.probs <- predict(enet_1se.weighted, newx=x_test_mat, type="response")
enet_1se.weighted.opt_treshold <- optimize_threshold(y_test, enet_1se.weighted.probs)

enet_1se.weighted.perf <- get_performances(
  y_true=y_test,
  y_probs=enet_1se.weighted.probs,
  threshold=enet_1se.weighted.opt_treshold$best_threshold
)
```

### EN WITH WEIGHTS - WITH ALL GENES

```{r}
enet_reg.genes.weighted <- glmnet(
  x_train_mat.genes, y_train.genes,
  alpha = enet.genes.weights.params$best_alpha,
  family="binomial", standardize=TRUE,
  weights = model_weights.genes
)

enet_cv_model.genes.weighted <- cv.glmnet(
  x_train_mat.genes, y_train.genes,
  alpha = enet.genes.weights.params$best_alpha,
  n_folds=10, family="binomial",
  standardize=TRUE,
  weights = model_weights.genes
)

enet_min_lambda.genes.weighted <- enet_cv_model.genes.weighted$lambda.min
enet_1se_lambda.genes.weighted <- enet_cv_model.genes.weighted$lambda.1se
```

```{r}
enet_min.genes.weighted <- glmnet(
  x_train_mat.genes, y_train.genes,
  alpha = enet.genes.weights.params$best_alpha,
  lambda = enet_min_lambda.genes.weighted,
  family="binomial", standardize=TRUE,
  weights = model_weights.genes
)

enet_1se.genes.weighted <- glmnet(
  x_train_mat.genes, y_train.genes,
  alpha = enet.genes.weights.params$best_alpha,
  lambda = enet_1se_lambda.genes.weighted,
  family="binomial", standardize=TRUE,
  weights = model_weights.genes
)
```

```{r}
enet_min.genes.weighted.probs <- predict(enet_min.genes.weighted, newx=x_test_mat.genes, type="response")
enet_min.genes.weighted.opt_treshold <- optimize_threshold(y_test.genes, enet_min.genes.weighted.probs)

enet_min.genes.weighted.perf <- get_performances(
  y_true=y_test.genes,
  y_probs=enet_min.genes.weighted.probs,
  threshold=enet_min.genes.weighted.opt_treshold$best_threshold
)

enet_1se.genes.weighted.probs <- predict(enet_1se.genes.weighted, newx=x_test_mat.genes, type="response")
enet_1se.genes.weighted.opt_treshold <- optimize_threshold(y_test.genes, enet_1se.genes.weighted.probs)

enet_1se.genes.weighted.perf <- get_performances(
  y_true=y_test.genes,
  y_probs=enet_1se.genes.weighted.probs,
  threshold=enet_1se.genes.weighted.opt_treshold$best_threshold
)
```

```{r}
# BASE EN - WITH CANCER SUBTYPE
cat("BASE EN - CANCER SUBTYPE\n")
cat("enet_min threshold:", enet_min.opt_treshold$best_threshold, "\n")
cat("enet_1se threshold:", enet_1se.opt_treshold$best_threshold, "\n")
cat("best alpha:", enet.params$best_alpha, "\n\n")

# BASE EN - WITH ALL GENES
cat("BASE EN - ALL GENES\n")
cat("enet_min.genes threshold:", enet_min.genes.opt_treshold$best_threshold, "\n")
cat("enet_1se.genes threshold:", enet_1se.genes.opt_treshold$best_threshold, "\n")
cat("best alpha:", enet.genes.params$best_alpha, "\n\n")

# EN UPSAMPLED - WITH ALL GENES
cat("EN UPSAMPLED - ALL GENES\n")
cat("enet_min.genes.up threshold:", enet_min.genes.up.opt_treshold$best_threshold, "\n")
cat("enet_1se.genes.up threshold:", enet_1se.genes.up.opt_treshold$best_threshold, "\n")
cat("best alpha:", enet.genes.up.params$best_alpha, "\n\n")

# EN UPSAMPLED - WITH CANCER SUBTYPE
cat("EN UPSAMPLED - CANCER SUBTYPE\n")
cat("enet_min.up threshold:", enet_min.up.opt_treshold$best_threshold, "\n")
cat("enet_1se.up threshold:", enet_1se.up.opt_treshold$best_threshold, "\n")
cat("best alpha:", enet.up.params$best_alpha, "\n\n")

# EN WITH WEIGHTS - WITH CANCER SUBTYPE
cat("EN WITH WEIGHTS - CANCER SUBTYPE\n")
cat("enet_min.weighted threshold:", enet_min.weighted.opt_treshold$best_threshold, "\n")
cat("enet_1se.weighted threshold:", enet_1se.weighted.opt_treshold$best_threshold, "\n")
cat("best alpha:", enet.weights.params$best_alpha, "\n\n")

# EN WITH WEIGHTS - WITH ALL GENES
cat("EN WITH WEIGHTS - ALL GENES\n")
cat("enet_min.genes.weighted threshold:", enet_min.genes.weighted.opt_treshold$best_threshold, "\n")
cat("enet_1se.genes.weighted threshold:", enet_1se.genes.weighted.opt_treshold$best_threshold, "\n")
cat("best alpha:", enet.genes.weights.params$best_alpha, "\n\n")

```

```{r}
# ----- FULL MODEL -----
full_opt_threshold <- optimize_threshold(y_test, full_model_probs)
full_perf <- get_performances(
  y_true = y_test,
  y_probs = full_model_probs,
  threshold = full_opt_threshold$best_threshold
)

# ----- STEPWISE AIC MODEL -----
stepwise_aic_opt_threshold <- optimize_threshold(y_test, stepwise_aic_probs)
stepwise_aic_perf <- get_performances(
  y_true = y_test,
  y_probs = stepwise_aic_probs,
  threshold = stepwise_aic_opt_threshold$best_threshold
)

# ----- STEPWISE BIC MODEL -----
stepwise_bic_opt_threshold <- optimize_threshold(y_test, stepwise_bic_probs)
stepwise_bic_perf <- get_performances(
  y_true = y_test,
  y_probs = stepwise_bic_probs,
  threshold = stepwise_bic_opt_threshold$best_threshold
)

# ----- Print all results -----
full_perf
stepwise_aic_perf
stepwise_bic_perf

```

## Accept Reject Lasso (ARL)

### With the clustered data

```{r}
library(stabs)
library(glmnet)

set.seed(123)

stab_lasso <- stabsel(
  x = x_train_mat,
  y = y_train,
  fitfun = glmnet.lasso, 
  cutoff = 0.65,      # selection probability threshold
  PFER = 10            # controls expected false selections
)

stable_features <- names(stab_lasso$selected)
length(stable_features)
```

```{r}
x_train_stable <- x_train_mat[, stable_features, drop = FALSE]
x_test_stable  <- x_test_mat[,  stable_features, drop = FALSE]

cv_stable_lasso <- cv.glmnet(
  x = x_train_stable,
  y = y_train,
  family = "binomial",
  alpha = 1
)

# Get fitted model
stable_lasso_model <- cv_stable_lasso$glmnet.fit

stable_probs <- predict(
  cv_stable_lasso,
  newx = x_test_stable,
  s = "lambda.min",
  type = "response"
)

stable_opt <- optimize_threshold(y_test, stable_probs)
stable_threshold <- stable_opt$best_threshold

stable_perf <- get_performances(
  y_true = y_test,
  y_probs = stable_probs,
  threshold = stable_threshold
)

stable_perf
```

### With all genes

```{r}
library(stabs)
library(glmnet)

set.seed(123)

stab_lasso.genes <- stabsel(
  x = x_train_mat.genes,
  y = y_train.genes,
  fitfun = glmnet.lasso, 
  cutoff = 0.6,      # selection probability threshold
  PFER = 50           # controls expected false selections
)

stable_features.genes <- names(stab_lasso.genes$selected)
length(stable_features.genes)
stable_features.genes
```

```{r}
x_train_stable.genes <- x_train_mat.genes[, stable_features.genes, drop = FALSE]
x_test_stable.genes  <- x_test_mat.genes[,  stable_features.genes, drop = FALSE]

cv_stable_lasso.genes <- cv.glmnet(
  x = x_train_stable.genes,
  y = y_train.genes,
  family = "binomial",
  alpha = 1,
  standardize = TRUE
)

# Get fitted model
lasso_stable.genes <- cv_stable_lasso.genes$glmnet.fit

stable_probs.genes <- predict(
  cv_stable_lasso.genes,
  newx = x_test_stable.genes,
  s = "lambda.min",
  type = "response"
)

stable_opt.genes <- optimize_threshold(y_test, stable_probs)
stable_threshold.genes <- stable_opt.genes$best_threshold

stable_perf.genes <- get_performances(
  y_true = y_test.genes,
  y_probs = stable_probs.genes,
  threshold = stable_threshold.genes
)

stable_perf.genes
```

## Variable decorrelation Lasso

```{r}
library(glmnet)
library(caret)
library(cluster)

cor_mat <- cor(x_train_mat.genes)
```

```{r}
d <- as.dist(1 - abs(cor_mat))   # strong correlations → small distance
clust <- hclust(d, method = "average")

# choose number of clusters automatically
k <- max(2, round(ncol(x_train_mat.genes) / 20))
clusters <- cutree(clust, k = k)
```

```{r}
safe_pca <- function(train_block, test_block) {
  # Force matrix form
  train_block <- as.matrix(train_block)
  test_block  <- as.matrix(test_block)
  
  if (ncol(train_block) == 1) {
    # If only one gene in the cluster → no PCA needed
    # Use standardized gene as the component
    mean_tr <- mean(train_block)
    sd_tr   <- sd(train_block)
    
    # Return scaled values
    return(list(
      train_pc1 = (train_block - mean_tr) / sd_tr,
      test_pc1  = (test_block - mean_tr) / sd_tr
    ))
  } else {
    # Standard PCA on clusters with >1 gene
    pc <- prcomp(train_block, scale = TRUE)
    return(list(
      train_pc1 = pc$x[, 1],
      test_pc1  = predict(pc, newdata = test_block)[, 1]
    ))
  }
}
```

```{r}
cluster_ids <- unique(clusters)

train_list <- list()
test_list  <- list()

for (cl in cluster_ids) {
  block_train <- x_train_mat.genes[, clusters == cl]
  block_test  <- x_test_mat.genes[,  clusters == cl]
  
  comp <- safe_pca(block_train, block_test)
  
  train_list[[as.character(cl)]] <- comp$train_pc1
  test_list[[as.character(cl)]]  <- comp$test_pc1
}

reduced_train <- do.call(cbind, train_list)
reduced_test  <- do.call(cbind, test_list)
```

```{r}
lasso_decorr <- cv.glmnet(
  x = reduced_train,
  y = y_train.genes,
  family = "binomial",
  alpha = 1
)

decorr_probs <- predict(lasso_decorr,
                        newx = reduced_test,
                        type = "response",
                        s = "lambda.min")

decorr_probs <- as.numeric(decorr_probs)

lasso_decorr_opt_treshold <- optimize_threshold(y_test.genes, decorr_probs)$best_threshold

lasso_decorr.perf <- get_performances(y_test.genes, decorr_probs, lasso_decorr_opt_treshold)

lasso_decorr.perf
```

## UniLasso

```{r}
get_univariate_pvals <- function(X, y) {
  pvals <- apply(X, 2, function(col) {
    model <- suppressWarnings(glm(y ~ col, family = "binomial"))
    summary(model)$coefficients[2, 4]  # p-value of predictor
  })
  return(pvals)
}

pvals <- get_univariate_pvals(x_train_mat.genes, y_train.genes)

K <- 30
top_genes <- names(pvals[pvals < 0.05])

length(top_genes)


Xtrain_uni <- x_train_mat.genes[, top_genes, drop = FALSE]
Xtest_uni  <- x_test_mat.genes[, top_genes, drop = FALSE]

lasso_uni <- cv.glmnet(
  x = Xtrain_uni,
  y = y_train.genes,
  family = "binomial",
  alpha = 1
)

uni_probs <- predict(lasso_uni,
                     newx = Xtest_uni,
                     type = "response",
                     s = "lambda.min")

uni_probs <- as.numeric(uni_probs)

uni_best_thresh <- optimize_threshold(y_test.genes, uni_probs)$best_threshold

uni.perf <- get_performances(y_test.genes, uni_probs, uni_best_thresh)
uni.perf
```
